A Parallel Algorithm for Solving Global Optimization Problems and Its Application for Tuning Hyperparameters of AI Methods

### 2025-06-18 2025/RuSCDays/RSD-2025.tex
some fix

%(рус.) Таким образом, возникает the hyperparameter optimization (HPO) problem, которая с математической точки зрения является задачей глобальной оптимизации. Типичным является случай наличия не только непрерывных, но и категориальных параметров, т.е. возникает необходимость использования методы глобальной оптимизации, способные решать задачи со смешанными параметрами.
Thus, there arises the hyperparameter optimization (HPO) problem, which from the mathematical point of view is a global optimization problem. In this case, it is typical to have not only continuous but also categorical parameters (parameters having a prefixed set of values), i.e. there is a need to use global optimization methods capable of solving problems with mixed parameters. 
Such methods belonging to the class of metaheuristic and model-based algorithms are implemented in the well-known hyperparameter optimization frameworks Optuna \cite{optuna} and HyperOpt \cite{hyperopt}.

@@

%
%
% !!!!!!!!!!!!!!! TO DO:: расширить обзор смежных исследований !!!!!!!!!!!!!
% "мало внимания обзору существующих методов и подходов для решения задачи оптимизации гиперпараметров"
%
%

@@

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig1.png}
\caption{Graph of the one-dimensional function after dimensionality reduction of multiextremal objective function.} \label{fig1}
\end{figure}

@@

%(рус.) На первой итерации алгоритма процесс-мастер (равномерно или случайно) выбирает p различных внутренних точек области поиска {x^1,x^2,…,x^p }, x^i∈(0,1),i=1,…,p, и рассылает точки испытаний по процессам-рабочим для выполнения испытаний в них. Правила работы процесса-мастера на любой из следующих итераций состоят в следующем. 
At the first iteration of the algorithm, the master process selects (uniformly or randomly) $p$ different internal points of the search domain ${x^1, x^2, ..., x^p}, x^i \in (0,1), i=1, ..., p$, and distributes the trial points to the worker processes to perform trials in those points. The rules for the master process at any of the subsequent iterations are as follows.

%(рус.) Пусть проведено k испытаний, их результаты получены и обработаны мастером, а на текущей итерации процессы-рабочие проводят испытания в точках множества S_k  = {x^(k+1),x^(k+2),…,x^(k+p) }, т.е. в этих точках испытания уже начались, но еще не завершены. Как только один из процессов-рабочих завершает проведение испытания в некоторой точке (например, x^(k+1)), он пересылает процессу-мастеру результаты испытания. Получив от рабочего процесса результаты испытания и обработав их мастер, в свою очередь, выбирает для него точку нового испытания x^(k+p+1) в соответствии со следующими правилами.
Suppose $k$ trials have been performed, the master has received and processed their results, and during the current iteration the worker processes perform trials at the points of the set $S_k={x^{k+1}, x^{k+2},..., x^{k+p}}$, i.e. the trials have already started at these points but have not yet been completed. As soon as one of the worker processes completes a trial at some point (e.g. $x^{k+1}$), it forwards the results of the trial to the master process and the point is excluded from the set $S_k$. After receiving the trial results from the worker process and processing them, the master process in turn selects a new trial point $x^{k+p+1}$ for the worker process according to the following rules.

\begin{enumerate}
%(рус.) 1. Перенумеровать в порядке возрастания все точки X_k  = {x^1,x^2,…,x^(k+p)  }, в которых либо проведены, либо проводятся испытания, т.е.
\item Renumber, in ascending order, all the points $X_k={x^1,x^2,...,x^{k+p} }$ at which trials have either been carried out or are being carried out, i.e.
\[
0 = x_1 \leq x_2 \leq ... \leq x_{k+p} = 1,
\]
%(рус.) и сопоставить им значения z_i=φ(y(x_i )) , 1≤i≤k, вычисленные в этих точках, а также индекс v_i=v(x_i ), определяемый по правилу
and assign to them the values $z_i = \phi(y(x_i)), 1 \leq i \leq k$ calculated at these points, as well as the index $v_i = v(x_i)$ determined by the rule

@@

%(рус.) 2. Вычислить значения
\item Calculate the values
\[
M_1= \max \left\{ \frac{|z_i - z_{i - 1}|}{(x_i - x_{i - 1})^\frac{1}{N}}: x_{i-1} \notin S_k, x_i \notin S_k, 2 \leq i \leq k+p \right\},
\]
\[
M_2= \max \left\{ \frac{|z_{i+1} - z_{i - 1}|}{(x_{i+1} - x_{i - 1})^\frac{1}{N}}: x_{i} \in S_k, 2 \leq i < k+p \right\},
\]
\[
M= \max \left\{ M_1, M_2 \right\},
\]
%(рус.) где z_i=f(x_i ), if x_i  ∉  S_k,1≤ i≤ k+p. Значения z_i в точках x_i∈S_k являются неопределенными, т.к. испытания в этих точках еще не завершены. Если значение M получилось равным 0, то положить M=1.
where $z_i=\phi(y(x_i))$, if $x_i \notin S_k, 1 \leq i \leq k+p$. The values $z_i$ at the points $x_i \in S_k$ are undefined because the trials at these points have not yet been completed. If the value $M$ turns out to be equal to $0$, then assume $M=1$.

@@

%(рус.) 4. Каждому интервалу (x_(i-1),x_i ),x_(i-1)  ∉ S_k,x_i  ∉ S_k,2≤ i≤ k+p, поставить в соответствие число R(i), которое называется характеристикой интервала и вычисляется по формуле
\item Assign a number $R(i)$ to each interval $(x_{i-1}, x_i), x_{i-1} \notin S_k, x_i \notin S_k, 2 \leq i \leq k+p$, which is called the interval characteristic and is calculated using the formula
\begin{equation}\label{characteristic}
R(i)=
\begin{cases}
    \Delta_i+\frac{(z_i-z_{i-1})^2}{(rM)^2\Delta_i} - 2 \frac{(z_i+z_{i-1}-2z^*)}{rM},   & \quad v(x_i)=v(x_{i-1})=1,\\
    2\Delta_i- 4 \frac{(z_i-z^*)}{rM},   & \quad v(x_i)=1, v(x_{i-1})=0,\\
    2\Delta_i- 4 \frac{(z_{i-1}-z^*)}{rM},       & \quad v(x_i)=0, v(x_{i-1})=1,\\
\end{cases}
\end{equation}
%(рус.) где Δ_i  =(x_i-x_(i-1) )^(1/N) , and r>1 – параметр алгоритма.
where $\Delta_i=(x_i-x_{i-1})^\frac{1}{N}$, and $r>1$ is the parameter of the algorithm.

%(рус.) 5. Выбрать интервал [x_(t-1),x_t ], которому соответствует максимальная характеристика, т.е.
\item Choose the interval $[x_{t-1}, x_t]$ to which the maximum characteristic corresponds, i.e.
\[
R(t)= \max \left\{ R(i): x_{i-1} \notin S_k, x_i \notin S_k, 2 \leq i \leq k+p\right\}.
\]

%(рус.) 6. Вычислить точку проведения нового испытаний x^{k+p+1} ∈ (x_(t-1),x_t ) в соответствии с формулами
\item Calculate the point of the new trial $x^{k+p+1} \in (x_{t-1}, x_t)$ according to the formulas
\begin{equation}\label{new_point_1} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}, \text{ if } v(x_{t-1}) \neq v(x_t),
\end{equation}
\begin{equation}\label{new_point_2} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}-\frac{\textit{sign}{(z_t-z_{t-1})}}{2r} \left[ \frac{|z_t-z_{t-1}|}{M} \right]^N,
\end{equation}
if $v(x_{t-1})=v(x_t)$.
%(рус.) Сразу после вычисления точки очередного испытания процесс-мастер добавляет ее в множество S и пересылает процессу-рабочему, который начинает проведение испытания в ней.
Immediately after calculating the point of the next trial, the master process adds it to the set $S$ and forwards it to the worker process, which starts a trial there.
\end{enumerate}

@@

%(рус.) Задачу со смешанными параметрами можно свести к решению взаимосвязанного набора задач с непрерывными параметрами. Занумеруем целыми числами все различные комбинации категориальных параметров, т.е. сопоставим каждому номеру s вектор u_s. Тогда (10) можно переписать в виде 
A problem with mixed parameters can be reduced to solving an interrelated set of problems with continuous parameters. Let us number all different combinations of categorical parameters with integers, i.e., let us assign to each number $s$ a vector $u_s$. Then (\ref{mixed_problem}) can be rewritten as

@@

%(рус.) Результаты проведенного эксперимента отражены в таблице 1: указано время решения задачи и значение целевой метрики, полученное с помощью iOpt и OPTUNA. Число точек, в которых не получилось вычислить критерий оптимизации, составило порядка 10\% от общего числа точек search trials. Кривые, соответствующие полученному предсказанию, приведены на Fig.2. Темная линия соответствует истинным значениям временного ряда, голубая – предсказанным значениям, полученным после использования OPTUNA, желтая линия – предсказанным значениям, полученным после использования iOpt.
%
%
%TO DO: "значение целевой метрики" - что за метрика использовалась в качестве целевой??? F_1????????
%
%
The results of the experiment are shown in Table~\ref{tab1}: the time required to solve the problem and the value of the objective metric $F_1$ obtained using iOpt and Optuna. The number of points where the optimization criterion could not be calculated was about 10\% of the total number of search trials. The curves corresponding to the obtained prediction are shown in Fig.~\ref{fig3}. The dark line corresponds to the true values of the time series, the blue line corresponds to the predicted values obtained after using Optuna, and the yellow line corresponds to the predicted values obtained after using iOpt.

@@

\begin{figure}
\center
\includegraphics[width=1\textwidth]{fig3.png}
\caption{Prediction visualization.} \label{fig3}
\end{figure}
%
%
%
%TO DO: указать единицу измерения времени (секунды?)
%
%
%
\begin{table}
\centering
\caption{Results of solving the problem using iOpt and Optuna.}\label{tab1}
\begin{tabular}{|l|c|c|c|c|}
\hline
\multirow{2}{*}{P} & \multicolumn{2}{c|}{iOpt} & \multicolumn{2}{c|}{Optuna} \\
\cline{2-5} 
 & Solving time & Final $F_1$-score & Solving time & Final $F_1$-score \\
\hline
1 & 6574.2 & 13.5 & 9554.1 & 13.4 \\
2 & 3579.2 & 13.9 & 5127.5 & 13.6 \\
4 & 1998.7 & 13.6 & 2598.6 & 13.2 \\
8 & 1331.7 & 13.2 & 1275.4 & 13.2 \\
16 & 846.6 & 13.3 & 1148.5 & 13.3 \\
20 & 699.9 & 13.9 & 1578.6 & 13.3 \\
\hline
\end{tabular}
\end{table}

>>>>

%(рус.) Таким образом, возникает the hyperparameter optimization (HPO) problem, которая с математической точки зрения является задачей глобальной оптимизации. Типичным является случай наличия не только непрерывных, но и категориальных параметров, т.е. возникает необходимость использования методы глобальной оптимизации, способные решать задачи со смешанными параметрами.
Thus, there arises the hyperparameter optimization (HPO) problem, which from the mathematical point of view is a global optimization problem. In this case, it is typical to have not only continuous but also categorical parameters (parameters having a prefixed set of values), i.e. there is a need to use global optimization methods capable of solving problems with mixed parameters. 
%new
Such methods belonging to the class of metaheuristic and model-based algorithms are implemented in the well-known hyperparameter optimization frameworks Optuna \cite{optuna} and HyperOpt \cite{hyperopt}.

@@

@@

\begin{figure}
\center
\includegraphics[width=0.8\textwidth]{fig1.png}
\caption{Graph of the one-dimensional function after dimensionality reduction of a multiextremal objective function.} \label{fig1}
\end{figure}

@@

%(рус.) На первой итерации алгоритма процесс-мастер (равномерно или случайно) выбирает p различных внутренних точек области поиска {x^1,x^2,…,x^p }, x^i∈(0,1),i=1,…,p, и рассылает точки испытаний по процессам-рабочим для выполнения испытаний в них. Правила работы процесса-мастера на любой из следующих итераций состоят в следующем. 
At the first iteration of the algorithm, the master process selects (uniformly or randomly) $p$ different internal points of the search domain $\{x^1, x^2, ..., x^p\}, x^i \in (0,1), 1 \leq i \leq p$, and distributes the trial points to the worker processes to perform trials in those points. The rules for the master process at any of the subsequent iterations are as follows.

%(рус.) Пусть проведено k испытаний, их результаты получены и обработаны мастером, а на текущей итерации процессы-рабочие проводят испытания в точках множества S_k  = {x^(k+1),x^(k+2),…,x^(k+p) }, т.е. в этих точках испытания уже начались, но еще не завершены. Как только один из процессов-рабочих завершает проведение испытания в некоторой точке (например, x^(k+1)), он пересылает процессу-мастеру результаты испытания. Получив от рабочего процесса результаты испытания и обработав их мастер, в свою очередь, выбирает для него точку нового испытания x^(k+p+1) в соответствии со следующими правилами.
Suppose $k$ trials have been performed, the master has received and processed their results, and during the current iteration the worker processes perform trials at the points of the set $W=\{x^{k+1}, x^{k+2},..., x^{k+p}\}$, i.e. the trials have already started at these points but have not yet been completed. As soon as one of the worker processes completes a trial at some point (e.g. $x^{k+1}$), it forwards the results of the trial to the master process and the point is excluded from the set $W$. After receiving the trial results from the worker process and processing them, the master process in turn selects a new trial point $x^{k+p+1}$ for the worker process according to the following rules.

\begin{enumerate}
%(рус.) 1. Перенумеровать в порядке возрастания все точки X_k  = {x^1,x^2,…,x^(k+p)  }, в которых либо проведены, либо проводятся испытания, т.е.
\item Renumber, in ascending order, all the points $X_k=\{x^1,x^2,...,x^{k+p} \}$ at which trials have either been carried out or are being carried out, i.e.
\[
0 = x_1 \leq x_2 \leq ... \leq x_{k+p} = 1,
\]
%(рус.) и сопоставить им значения z_i=φ(y(x_i )) , 1≤i≤k, вычисленные в этих точках, а также индекс v_i=v(x_i ), определяемый по правилу
and assign to them the values $z_i = \phi(y(x_i)), 1 \leq i \leq k$ calculated at these points, as well as the index $v_i = v(x_i)$ determined by the rule

@@

%(рус.) 2. Вычислить значения
\item Calculate the values
\[
M_1= \max \left\{ \frac{|z_i - z_{i - 1}|}{(x_i - x_{i - 1})^\frac{1}{N}}: x_{i-1} \notin W, x_i \notin W, 2 \leq i \leq k+p \right\},
\]
\[
M_2= \max \left\{ \frac{|z_{i+1} - z_{i - 1}|}{(x_{i+1} - x_{i - 1})^\frac{1}{N}}: x_{i} \in W, 2 \leq i < k+p \right\},
\]
\[
M= \max \left\{ M_1, M_2 \right\},
\]
%(рус.) где z_i=f(x_i ), if x_i  ∉  S_k,1≤ i≤ k+p. Значения z_i в точках x_i∈S_k являются неопределенными, т.к. испытания в этих точках еще не завершены. Если значение M получилось равным 0, то положить M=1.
where $z_i=\phi(y(x_i))$, if $x_i \notin W, 1 \leq i \leq k+p$. The values $z_i$ at the points $x_i \in W$ are undefined because the trials at these points have not yet been completed. If the value $M$ turns out to be equal to $0$, then assume $M=1$.

@@

%(рус.) 4. Каждому интервалу (x_(i-1),x_i ),x_(i-1)  ∉ S_k,x_i  ∉ S_k,2≤ i≤ k+p, поставить в соответствие число R(i), которое называется характеристикой интервала и вычисляется по формуле
\item Assign a number $R(i)$ to each interval $(x_{i-1}, x_i), x_{i-1} \notin W, x_i \notin W, 2 \leq i \leq k+p$, which is called the interval characteristic and is calculated using the formula
\begin{equation}\label{characteristic}
R(i)=
\begin{cases}
    \Delta_i+\frac{(z_i-z_{i-1})^2}{(rM)^2\Delta_i} - 2 \frac{(z_i+z_{i-1}-2z^*)}{rM},   & \quad v(x_i)=v(x_{i-1})=1,\\
    2\Delta_i- 4 \frac{(z_i-z^*)}{rM},   & \quad v(x_i)=1, v(x_{i-1})=0,\\
    2\Delta_i- 4 \frac{(z_{i-1}-z^*)}{rM},       & \quad v(x_i)=0, v(x_{i-1})=1,\\
\end{cases}
\end{equation}
%(рус.) где Δ_i  =(x_i-x_(i-1) )^(1/N) , and r>1 – параметр алгоритма.
where $\Delta_i=(x_i-x_{i-1})^\frac{1}{N}$, and $r>1$ is the parameter of the algorithm.

%(рус.) 5. Выбрать интервал [x_(t-1),x_t ], которому соответствует максимальная характеристика, т.е.
\item Choose the interval $[x_{t-1}, x_t]$ to which the maximum characteristic corresponds, i.e.
\[
R(t)= \max \left\{ R(i): x_{i-1} \notin W, x_i \notin W, 2 \leq i \leq k+p\right\}.
\]

%(рус.) 6. Вычислить точку проведения нового испытаний x^{k+p+1} ∈ (x_(t-1),x_t ) в соответствии с формулами
\item Calculate the point of the new trial $x^{k+p+1} \in (x_{t-1}, x_t)$ according to the formulas
\begin{equation}\label{new_point_1} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}, \text{ if } v(x_{t-1}) \neq v(x_t),
\end{equation}
\begin{equation}\label{new_point_2} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}-\frac{\textit{sign}{(z_t-z_{t-1})}}{2r} \left[ \frac{|z_t-z_{t-1}|}{M} \right]^N,
\end{equation}
if $v(x_{t-1})=v(x_t)$.
%(рус.) Сразу после вычисления точки очередного испытания процесс-мастер добавляет ее в множество S и пересылает процессу-рабочему, который начинает проведение испытания в ней.
Immediately after calculating the point of the next trial, the master process adds it to the set $W$ and forwards it to the worker process, which starts a trial there.
\end{enumerate}

@@

%(рус.) Задачу со смешанными параметрами можно свести к решению взаимосвязанного набора задач с непрерывными параметрами. Занумеруем целыми числами все различные комбинации категориальных параметров, т.е. сопоставим каждому номеру s вектор u_s. Тогда (10) можно переписать в виде 
A problem with mixed parameters can be reduced to solving an interrelated set of problems with continuous parameters. Let us number all different combinations of categorical parameters with integers, i.e., let us assign to each number $s \in \{1, S\}$ a vector $u_s$. Then (\ref{mixed_problem}) can be rewritten as

@@

%(рус.) Результаты проведенного эксперимента отражены в таблице 1: указано время решения задачи и значение целевой метрики, полученное с помощью iOpt и OPTUNA. Число точек, в которых не получилось вычислить критерий оптимизации, составило порядка 10\% от общего числа точек search trials. Кривые, соответствующие полученному предсказанию, приведены на Fig.2. Темная линия соответствует истинным значениям временного ряда, голубая – предсказанным значениям, полученным после использования OPTUNA, желтая линия – предсказанным значениям, полученным после использования iOpt.
%
%
%TO DO: "значение целевой метрики" - что за метрика использовалась в качестве целевой??? F_1????????
%
%
The results of the experiment are shown in Table~\ref{tab1}: the time required to solve the problem and the value of the objective metric MSE obtained using iOpt and Optuna. The number of points where the optimization criterion could not be calculated was about 10\% of the total number of search trials. The curves corresponding to the obtained prediction are shown in Fig.~\ref{fig3}. The dark line corresponds to the true values of the time series, the blue line corresponds to the predicted values obtained after using Optuna, and the yellow line corresponds to the predicted values obtained after using iOpt.

@@

\begin{figure}
\center
\includegraphics[width=1\textwidth]{fig3.png}
\caption{Prediction visualization.} \label{fig3}
\end{figure}
%
%
%
%TO DO: указать единицу измерения времени (секунды?)
%
%
%
\begin{table}
\centering
\caption{Results of solving the problem using iOpt and Optuna.}\label{tab1}
\begin{tabular}{|l|c|c|c|c|}
\hline
\multirow{2}{*}{P} & \multicolumn{2}{c|}{iOpt} & \multicolumn{2}{c|}{Optuna} \\
\cline{2-5} 
 & Solving time, sec. & Final MSE & Solving time, sec. & Final MSE \\
\hline
1 & 6574.2 & 13.5 & 9554.1 & 13.4 \\
2 & 3579.2 & 13.9 & 5127.5 & 13.6 \\
4 & 1998.7 & 13.6 & 2598.6 & 13.2 \\
8 & 1331.7 & 13.2 & 1275.4 & 13.2 \\
16 & 846.6 & 13.3 & 1148.5 & 13.3 \\
20 & 699.9 & 13.9 & 1578.6 & 13.3 \\
\hline
\end{tabular}
\end{table}
